|

### 2023-09-06 manuscript/manuscript_1.0.Rmd
Addressing DAniel´s comments

---
title: |
  The role of learning in status signalling: a modeling approach
author:
- Andrés E. Quiñones:
    email: "andresqp@gmail.com"
    institute: [andes, unine]
    correspondence: true
- Redouan Bshary:
    institute: unine
- C. Daniel Cadena:
    institute: andes
institute:
- andes: Laboratorio de Biología Evolutiva de Vertebrados, Departamento  de Ciencias Biológicas, Universidad de los Andes, Bogotá, Colombia
- unine: Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  # word_document: default
  # pdf_document:
    latex_engine: xelatex # default  #
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
fig_caption: yes
fontsize: 12pt
header-includes: |
  ```{=latex}
    \usepackage{float}
    \usepackage{unicode-math}
    \usepackage{setspace}\doublespacing
    \usepackage[backref=page]{hyperref}
    \linespread{2}
    \usepackage{lineno}
    \linenumbers
    \floatplacement{figure}{H}
    \newcommand{\beginsupplement}{ \setcounter{table}{0}         \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  ```
keep_tex: yes
keywords: My keywords
linespread: 2
bibliography: ../SocialMorphCue.bib
spacing: double
style: C:\Users\andre\Zotero\styles\elife.csl
abstract: |
  Adaptive behavioural responses often depend on qualities of the interacting
  partner of an individual. For example, when competing for resources, an individual
  might be better off escalating fights with individuals of lower quality, while
  restraining  from fighting individuals of higher quality. Communication systems
  involving signals of quality allow individuals to reduce uncertainty regarding the
  fighting ability of their partners and make more adaptive behavioral decisions. 
  However, dishonest individuals can destabilize such communications systems. 
  An open question is whether cognitive mechanisms, such as learning, 
  can maintain the honesty of signals, thus favoring their evolutionary 
  stability. Here we present evolutionary simulations where individuals 
  can produce a signal proportional to their quality and learn along their
  lifetime the best response to the signals emitted by their peers. Our simulations
  replicate previous results where the handicap principle mediates the evolution 
  of signals. As in traditional models without learning, in our simulations 
  learning on the receiver side can mediate the evolution of signals
  of quality on the sender side. When the cost of the signal is proportional to the
  quality of the sender, populations are only composed of honest signalers. In contrast 
  to traditional models which predict the absence of signals when the
  cost is not proportional to the quality of the signaler, our model revealed that
  learning facilitates the evolution of a polymorphism in which populations are
  composed of both honest and dishonest signalers. We argue that learning can 
  be a general cognitive mechanism having a role in a wide range of communication systems 
  and more generally in behavioral responses.
---

@@

The outcome of interactions with conspecific individuals is an important 
determinant of fitness in social animals. Irrespective of whether 
such interactions are cooperative or competitive, the actions of 
interacting individuals partly determine 
their reproductive success. However, the best action for an individual 
engaging in a social interaction might vary depending on its 
own condition, and that of individuals with 
whom it interacts. Thus, acting based on information about interacting 
partners is typically adaptive [@quinones_Negotiation_2016], yet
acquiring such information is far from trivial. 
In some cases, interacting partners (*i.e.* signallers) might be 'willing' to 
provide accurate information, but in others it might be in their 
own interest to conceal information [@johnstone_Recognition_1997] 
or to provide deceptive information [@johnstone_Badges_1993]. 
Typically, in a given context, some individuals benefit 
from broadcasting accurate information, whereas others benefit 
from concealing it. Take, for instance, an interaction between two individuals 
where one can help the other. Given some costs and benefits, 
a potential donor is likely to be interested in helping 
related individuals. Therefore, for a relative of the donor, 
broadcasting kinship would be advantageous, whereas concealing the lack 
of kinship would be better for an unrelated individual. 
Similar scenarios may apply to a variety of aspects of social 
life such as finding mates, feeding offspring, or engaging in
dominance relationships or aggressive interactions 
[@bradbury_Principles_2011; @moller_Female_1988; @tibbetts_Socially_2004].

In dominance and aggressive interactions, a crucial piece of
information to guide individual actions is the fighting ability of interacting 
partners, which is sometimes referred to as Resource Holding Potential 
[RHP, Parker -@parker_Assessment_1974], or simply as quality. 
Responsiveness to the quality of an opponent is central 
to communication systems such as those in which individuals 
exhibit badges of status, where a signal conveys quality 
[@johnstone_Badges_1993; @rohwer_Social_1975a]. 
Such signals must be arbitrary, meaning the signals could potentially be 
produced by high and low quality individuals alike. 
Generally signals can be evolutionarily stable whenever low quality individuals 
pay a higher fitness cost for carrying the signal such that it is not 
in the interest of low quality individuals to fake quality by producing a large badge.
Signal costs inversely proportional to quality may come about as a result of the
production costs of the signal, in which case the signal works like a handicap
[@botero_Evolution_2010; @johnstone_Badges_1993;@grafen_Biological_1990;@zahavi_Mate_1975]. 
Alternatively, the costs can come from 
the aggressive reaction of receivers when the convention of the communication 
is broken [@enquist_Signaling_2010;@tibbetts_Socially_2004]. For example in paper 
wasps, subordinate individuals with experimentally manipulated dominant-like facial
patterns received more aggression from the dominant. In this case, the costs are 
socially mediated rather than developmentally and are triggered by the violation 
of the convention. It is still incluer under this narrative, how the convenvention 
is stablished in the first place.

@@

<!-- Continuar aquí en la revisión -->

Associative learning is a taxonomically widespread cognitive 
mechanism that allows individuals to associate rewards with 
environmental stimuli and thus behave adaptively 
[@heyes_Simple_2012;@macphail_Brain_1982; @staddon_Adaptive_2016; 
@behrens_Associative_2008]. Theory has shown that natural selection 
favours such associations in complex environments where conditions 
are difficult to predict [@dridi_Environmental_2016]. 
Associative learning is a flexible cognitive 
mechanism whose underpinnings show interspecific variation 
[@enquist_Power_2016; @quinones_Reinforcement_2020;@pretot_Factors_2016;
pretot_Comparative_2021]. For example, species can have different learning rates
[@hoedjes_Natural_2010],
some may vary these rates during learning trials which allows them to have more 
flexible learning [@leimar_Flexibility_2023a], and some may include future rewards in their associations [raby_Planning_2007;osvath_Spontaneous_2009].
Associative learning is not a single mechanism, but rather a set of 
cognitive structures and processes that can vary in their scope and complexity. Presumably, these structures and processes have been modified 
by natural selection and could provide new explanations for behavioural 
variation. Nonetheless, associative learning is 
not often included in the narrative of evolutionary explanations of 
behavioural patterns [@fawcett_Exposing_2013; @kamil_Optimal_1983; 
@mcauliffe_Psychology_2015]. Computational models of evolution 
can play an important role to overcome the lack of integration 
between learning and evolution. Reinforcement learning theory 
encompasses a series of computational methods inspired 
on the psychological and neurological mechanisms of associative 
learning [@sutton_Reinforcement_2018]. This set of algorithms allows the 
implementation of biologically realistic problems, 
capturing the essence of learning processes 
[@frankenhuis_Enriching_2018; @quinones_Reinforcement_2020]. 
Furthermore, these algorithms can be embedded in evolutionary simulations to 
generate theoretical predictions of the effect that learning can 
have in behavioural evolution [@leimar_Learning_2019;@leimar_Effects_2022].

@@

The decision of whether to escalate a fight against an interacting 
partner can depend on the size of the partner's badge, 
the dependency is determined by the focal's experiences acquired
through a learning process. We implement learning using 
the actor-critic approach from reinforcement learning (RL) theory [@sutton_Reinforcement_2018;@quinones_Reinforcement_2020; @leimar_Learning_2019;
@leimar_Evolution_2021].
Individuals estimate the reward (pay-off) expected from  interacting 
with partners of different badge sizes (the critic in RL terminology). 
After each interaction they update the estimate of reward proportionally 
to the difference between their current estimate and the observed 
reward (prediction error $\delta$) and to the speed of learning 
($\mathrm{A}$). Furthermore, individuals express different probabilities of 
retreating/attacking depending on the badge size of their opponent 
(the actor in RL). They update the probability of 
retreating/attacking depending on whether retreating/attacking
leads to an increase in the reward estimation. 
So, if a focal individual decides to escalate a fight against 
an individual with a small badge and this leads to an increase in 
the reward estimation, then the focal individual will increase the probability 
of escalating fights with individuals of small badges in the future. 
Given that badge size is a real number between 0 and 1, there are infinitely 
many badge sizes. Thus, the reward estimation, as well as the probability 
of retreating/attacking, must be generalized across different values. 
To implement generalization we use the linear function approximation 
method based on radial basis functions [@sutton_Reinforcement_2018]. 
Specifically, we pick $c$ feature centres, which are evenly spaced values 
along the badge size interval ([0,1]) where the updates are focused. 
For the sake of simplicity we keep the location of these feature centres 
constant and  the same for all individuals, the are stored in the 
vector $\symbf{b}$ . Each of these feature centres 
is associated with a weight for reward estimation and tendency to play 
retreat in a given interaction. The reward estimation and the tendency 
to retreat are calculated (in every interaction) as the sum of the weights 
associated with each feature centre, weighted by the response 
triggered by the feature (Fig. \ref{fig:model-struc} B, dots 
represent the feature weights). The response of each feature centre 
diminishes as a Gaussian function with the distance between the 
feature centre and the badge size of 
the partner (Fig. \ref{fig:model-struc} B, grey line). Formally, the 
reward estimation $\hat{R}$ when the focal individual ($i$) faces 
individual $j$ is given by,
\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_i-b_z|}{2\theta^2})} 
\end{equation}

@@

```{r model-struc,echo=FALSE,warning=FALSE,fig.width=8,fig.align='center',fig.height=7,fig.show='hold',fig.cap= "Model of communication in the context of aggressive interactions. In A, the reaction norm determines the bib (badge) size, black band on birds chest. Red shows an informative reaction norm, whereas blue shows a uninformative reaction norm. B individuals perceiving the signal have a behavioral reaction norm that determines their probability of retreating (black line). The reaction norm arises from generalizing the information from the feature weights (black dots). Feature weaights are values associated with a particular badge size and are updated as individuals interact with each other. The generalization of the information associated with each feature weight is represented by the grey line and its axis, which shows that the response triggered by the fourth feature diminishes as the value evaluated is further from the feature center. The learning process moves the feature weights (black dots) up or down depending on whether this leads to an increase in the estimated reward. In C and D, effect of learning on the receiver strategy. Receivers in C face signallers with uninformative reaction norms (blue line in A). The uninformative reaction norm yields a badge size of 0.5, thus the behavioural reaction norm in C is only updated around that value. D Receivers face signallers with informative reaction norms (red line in A). Accordingly, receivers develop a threashold-like reaction norm where the decision to retreat increases with the badge size of the interacting partner. Colour scale in C and D indicates the quality of the individual. "}

@@

We first present the outcome of simulations where we prevented the 
evolution of the badge size (by setting the mutation rate to $0$), 
and assumed that all individuals in the population display either 
an uninformative or informative badge size with respect to their quality. 
These simulations show what type of receiver strategy develops through 
a learning process in such monomorphic population. 
Figure \ref{fig:model-struc} (C and D) shows 
the receiver strategy developed through learning; panel C is for 
receivers that faced uninformative signals, and D informative ones. 
The simulations reveal that when learners face uninformative signals (Fig. \ref{fig:model-struc} C), they modify their probability of retreating 
depending on their own quality. 
Individuals with higher quality (red tones) have a 
high probability of attacking after the learning process; 
while individuals of lower quality 
(blue tones) mostly retreat from confrontations; in both cases 
individuals behave equally irrespective of the quality of
their opponent. Thus, learning splits the population of receivers into 
the two classic pure strategies of hawks and doves. 
Given that we have a assumed a monomorphic population 
with unresponsive reaction norms on the signalling side, the changes 
triggered by learning only affect a small range of badge sizes. Specifically, 
all learning has occurred around a badge size of $0.5$, because all
badges in the population are of this size (panel C). 
In contrast, when receivers face informative reaction norms 
on the side of the signaller (Fig. \ref{fig:model-struc} B), 
receivers use information on the badge size of their interacting 
partners to determine whether to retreat or attack. 
The relationship is given by a threshold-like reaction norm, 
where the decision to retreat or to attack depends on the quality of the receiver. 
As expected, the higher the quality of the receiver, the larger the badge size 
of the signaller that triggers a retreat.

@@

When we allowed the badge size to evolve ($\alpha$ and $\beta$ 
change subject to natural selection and genetic drift), and the signal
worked as a handicap (*i.e.* the cost of the badge is inversely 
proportional to the quality of the individual), the sender code evolved often
to produce an honest signal (Fig. \ref{fig:handicap}). The evolution of 
the sender code did not happen immediately after the start of the 
evolutionary process. Instead, the evolutionary dynamics of the reaction 
norm parameters (Fig. \ref{fig:handicap}, A) appeared to involve a set of 
steps. First, the intercept of the reaction norm ($\alpha$) 
describing the relationship between individual quality and badge size evolved to 
higher values, reducing the average badge size in the populations. 
This is because reaction norms segregating in the population are 
flat at the beginning of the simulation; consequently, 
the receiver code does not respond to the badge size 
and larger badges are costlier and do not trigger lower attack 
probabilities. During the first generations, the slope of the sender 
reaction norm ($\beta$)  remains close to the initial value of zero. 
At around generation 4000, the slope evolves toward positive values. 
Positive values in the slope imply that larger badges correlate 
with higher quality (Fig. \ref{fig:handicap} B). Receivers then learn to react to such 
correlation, increasing the probability of retreating when facing 
individuals with larger badges (Fig. \ref{fig:handicap} C). Hence, natural selection 
favours larger values of $\beta$, eventually leading to an evolutionary 
equilibrium in which badge size is an honest signal of quality mediated
by the learned responses of receivers (Fig. \ref{fig:handicap} B.4 and C.4).

@@

The amount of information that agents are able to collect through learning 
throughout their life time can strongly change the outcome of 
evolutionary dynamics. In the simulations presented in figure 
\ref{fig:handicap} and \ref{fig:noBadge}, 
individuals learned with high speed of learning ($\mathrm{A} = 0.4$) and interacted 
repeatedly along their lifetime (2000 interactions on average). When we reduced 
the number of interactions that individuals experienced over their lifetime to 300 
on average, we saw a drastic increase in the phenotypic and genetic
variation present in the population. Populations start monomorphic 
with a value of zero on both the intercept and
the slope, and mutations quickly build up a normal distribution around the starting 
value. Within the first 2000 generations, the unimodal distribution in 
the intercept ($\alpha$) splits into a bimodal one. 
Later in evolutionary time one of the peaks splits further, 
so at the end of the evolutionary simulation the distribution of the intercept 
($\alpha$) in the population shows three peaks. In the case of $\beta$, the 
peaks in the distribution are not so clear-cut, but there is an 
increase in the variance of the distribution over time. These changes in the 
distribution of the parameters of the sender reaction norm 
imply that individuals can generally be classified into three 
distinct types (Fig. \ref{fig:branching}). Two types express a flat reaction 
norm with extreme values for the badge size, meaning that their badge size 
is not informative of their quality, whereas the third type shows 
intermediate badge sizes determined by the quality of the individual 
(Fig. \ref{fig:branching} B.3-4). 
In this simulation, the values of the slope in the informative group
are negative, which implies that badge size correlates negatively with quality. 
Furthermore, the receiver reaction norms developed through learning, particularly those 
of individuals of intermediate quality, respond to the signal of their 
sender type by increasing the probability of retreating from a fight with 
individuals with smaller badges (Fig. \ref{fig:branching}).

@@

A limited number of interactions has an effect on the amount of variation in
the evolving parameters when the signal follows the handicap
principle as well. In fig \ref{fig:honestDiv}, we show simulations where individuals 
have on average 300 interactions in their life and the cost of the signal
is proportional to the quality, following the handicap principle. 
The evolutionary process leads to a combination of reaction norm parameters where 
there is a positive correlation between the size of the badge and the quality 
of the individual. This relationship, however, is muddled by the fact that there 
are two clusters of values for the intercept ($\alpha$) and slope ($\beta$) 
of the reaction norm in the population. Thus, there are two types 
of reaction norms. One of the types has a steeper slope compared to the other, 
meaning that it expresses a higher badge size for a given quality. 
This effect of an increased variance in the trait distribution
is not only triggered by lower number of interactions. 
Larger variances are also found when we assume 
a lower speed of learning (data not shown). This suggest
that limits to the amount of information that individuals acquire through
learning allows the coexistence of different communication strategies within 
a population, something previously shown by @botero_Evolution_2010.

@@

The action of natural selection and genetic drift on the learning parameters 
do not override the conditions for the evolutionary branching 
[@dieckmann_Dynamical_1996] described earlier. 
Given how crucial naive behaviour is for the diversification process 
described earlier, we tested whether evolutionary processes could lead 
key parameters of the learning process toward
values that prevented the genetic variation to build up. 
Figure \ref{fig:learnEvol} shows the dynamics from a set of simulations where we 
allow for the coevolution of the reaction norm as well as key parameters of
the learning module, namely the speed of learning ($A$) and the behavioural 
tendency expressed by individuals before learning. Top panels in 
figure \ref{fig:learnEvol} show that there is diversification process in this 
coevolutionary scenario as well. That is because, as is shown in the middle 
panels (B) of the same figure, the learning parameters stay within the range 
necessary for the diversification process. The speed of learning 
(Fig. \ref{fig:learnEvol} B.1) stays well above zero throughout the simulation. 
In contrast, the initial behavioral tendency maintain values around zero. 
Thus, evolutionary processes acting on the learning parameters 
did not override the conditions necessary for frequency dependent to boost 
genetic variation.

>>>>

---
title: |
  The role of learning in status signalling: a modeling approach
author:
- Andrés E. Quiñones:
    email: "andresqp@gmail.com"
    institute: [andes, unine]
    correspondence: true
- Redouan Bshary:
    institute: unine
- C. Daniel Cadena:
    institute: andes
institute:
- andes: Laboratorio de Biología Evolutiva de Vertebrados, Departamento  de Ciencias Biológicas, Universidad de los Andes, Bogotá, Colombia
- unine: Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex  # default  ##word_document: default  # word_document:default
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  word_document: default
fig_caption: yes
fontsize: 12pt
header-includes: |
  ```{=latex}
    \usepackage{float}
    \usepackage{unicode-math}
    \usepackage{setspace}\doublespacing
    \usepackage[backref=page]{hyperref}
    \linespread{2}
    \usepackage{lineno}
    \linenumbers
    \floatplacement{figure}{H}
    \newcommand{\beginsupplement}{ \setcounter{table}{0}         \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  ```
keep_tex: yes
keywords: My keywords
linespread: 2
bibliography: ../SocialMorphCue.bib
spacing: double
style: C:\Users\andre\Zotero\styles\elife.csl
abstract: |
  Adaptive behavioural responses often depend on qualities of the interacting
  partner of an individual. For example, when competing for resources, an individual
  might be better off escalating fights with individuals of lower quality, while
  restraining  from fighting individuals of higher quality. Communication systems
  involving signals of quality allow individuals to reduce uncertainty regarding the
  fighting ability of their partners and make more adaptive behavioral decisions. 
  However, dishonest individuals can destabilize such communications systems. 
  An open question is whether cognitive mechanisms, such as learning, 
  can maintain the honesty of signals, thus favoring their evolutionary 
  stability. We present evolutionary simulations where individuals 
  can produce a signal proportional to their quality and learn along their
  lifetime the best response to the signals emitted by their peers. Our simulations
  replicate previous results where the handicap principle mediates the evolution 
  of signals. In our simulations learning on the receiver side can mediate the evolution
  of signals  of quality on the sender side. When the cost of the signal is proportional
  to the  quality of the sender, all individuals in populations are honest signalers.   
  In contrast to traditional models which predict the absence of signals when the
  cost is not proportional to the quality of the signaler, our model revealed that
  learning facilitates the evolution of a polymorphism in which populations comprise both
  honest and dishonest signalers. We argue that learning may have a role in the evolution
  and dynamics of a wide range of communication systems and more generally in 
  behavioral responses.
---

@@

The outcome of interactions with conspecific individuals is a critical
determinant of fitness in social animals. Irrespective of whether 
such interactions are cooperative or competitive, the actions of 
interacting individuals influence 
their reproductive success. However, the best action for an individual 
engaging in a social interaction might vary depending on its 
own condition, and that of individuals with 
whom it interacts. Thus, acting based on information about interacting 
partners is typically adaptive [@quinones_Negotiation_2016], yet
acquiring such information is far from trivial. 
In some cases, interacting partners (*i.e.* signallers) might be 'willing' to 
provide accurate information, but in others it might be in their 
own interest to conceal information [@johnstone_Recognition_1997] 
or to provide deceptive information [@johnstone_Badges_1993]. 
Typically, in a given context, some individuals benefit 
from broadcasting accurate information, whereas others benefit 
from concealing it. Take, for instance, an interaction between two individuals 
where one can help the other. Given some costs and benefits, 
a potential donor is likely to be interested in helping 
related individuals. Therefore, for a relative of the donor, 
broadcasting kinship would be advantageous, whereas concealing the lack 
of kinship would be better for an unrelated individual. 
Similar scenarios may apply to a variety of aspects of social 
life such as finding mates, feeding offspring, or engaging in
dominance relationships or aggressive interactions 
[@bradbury_Principles_2011; @moller_Female_1988; @tibbetts_Socially_2004].

In dominance and aggressive interactions, a crucial piece of
information to guide individual actions is the fighting ability of interacting 
partners, which is sometimes referred to as Resource Holding Potential 
[RHP, Parker -@parker_Assessment_1974] or simply as quality. 
Responsiveness to the quality of an opponent is central 
to communication systems in which individuals 
exhibit badges of status, where a signal conveys quality 
[@johnstone_Badges_1993; @rohwer_Social_1975a]. 
Such signals must be arbitrary, meaning the signals could potentially be 
produced by high and low quality individuals alike. 
Generally signals can be evolutionarily stable whenever low-quality individuals 
pay a higher fitness cost for carrying the signal such that it is not 
in the interest of low-quality individuals to fake quality by producing a large badge.
Signal costs inversely proportional to quality may come about as a result of the
production costs of the signal, in which case the signal works like a handicap
[@botero_Evolution_2010; @johnstone_Badges_1993;@grafen_Biological_1990;@zahavi_Mate_1975]. 
Alternatively, the costs can come from 
the aggressive reaction of receivers when the convention of the communication 
is broken [@enquist_Signaling_2010;@tibbetts_Socially_2004]. For example in paper 
wasps, subordinate individuals with experimentally manipulated dominant-like facial
patterns received more aggression from the dominant[@tibbetts_Socially_2004]. 
In this case, costs are socially rather than developmentally mediated and are 
triggered by the violation of the convention, although how the convention
is established in the first place is unclear.

@@

Associative learning is a taxonomically widespread cognitive 
mechanism that allows individuals to associate rewards with 
environmental stimuli and thus behave adaptively 
[@heyes_Simple_2012;@macphail_Brain_1982; @staddon_Adaptive_2016; 
@behrens_Associative_2008]. Theory has shown that natural selection 
favours such associations in complex environments where conditions 
are difficult to predict [@dridi_Environmental_2016]. 
Associative learning is a flexible cognitive 
mechanism whose underpinnings show interspecific variation 
[@enquist_Power_2016; @quinones_Reinforcement_2020;@pretot_Factors_2016;
@pretot_Comparative_2021]. For example, species can have different learning rates
[@hoedjes_Natural_2010], some may vary these rates during learning 
trials which allows them to have more 
flexible learning [@leimar_Flexibility_2023], and some may include future rewards in their associations [@raby_Planning_2007;@osvath_Spontaneous_2009].
Associative learning is not a single mechanism, but rather a set of 
cognitive structures and processes that can vary in their scope and complexity. Presumably, these structures and processes have been modified 
by natural selection and could provide novel explanations for behavioural 
variation. Nonetheless, associative learning is 
not often included in the narrative of evolutionary explanations of 
behavioural patterns [@fawcett_Exposing_2013; @kamil_Optimal_1983; 
@mcauliffe_Psychology_2015].

Computational models of evolution can overcome the lack of integration 
between learning and evolution. Reinforcement 
learning theory encompasses a series of computational methods inspired 
on the psychological and neurological mechanisms of associative 
learning [@sutton_Reinforcement_2018]. This set of algorithms allows the 
implementation of biologically realistic problems, 
capturing the essence of learning processes 
[@frankenhuis_Enriching_2018; @quinones_Reinforcement_2020]. 
Furthermore, these algorithms can be embedded in evolutionary simulations to 
generate theoretical predictions of the effect of learning on behavioural evolution [@leimar_Learning_2019;@leimar_Effects_2022].

@@

The decision of whether to escalate a fight against an interacting 
partner can depend on the size of the partner's badge, 
with the dependency being determined by the focal's experiences acquired
through a learning process. We implement learning using 
the actor-critic approach from reinforcement learning (RL) theory [@sutton_Reinforcement_2018;@quinones_Reinforcement_2020; @leimar_Learning_2019;
@leimar_Evolution_2021].
Individuals estimate the reward (pay-off) expected from  interacting 
with partners of different badge sizes (the critic in RL terminology). 
After each interaction they update the estimate of reward proportionally 
to the difference between their current estimate and the observed 
reward (prediction error $\delta$) and to the speed of learning 
($\mathrm{A}$). Furthermore, individuals express different probabilities of 
retreating/attacking depending on the badge size of their opponent 
(the actor in RL). They update the probability of 
retreating/attacking depending on whether retreating/attacking
leads to an increase in the reward estimation. 
Thus, if a focal individual decides to escalate a fight against 
an individual with a small badge and this leads to an increase in 
the reward estimation, then the focal individual will increase the probability 
of escalating fights with individuals of small badges in the future. 
Given that badge size is a real number between 0 and 1, there are infinitely 
many badge sizes. Thus, the reward estimation, as well as the probability 
of retreating/attacking, must be generalized across different values. 
To implement generalization we use the linear function approximation 
method based on radial basis functions [@sutton_Reinforcement_2018].
Specifically, we pick $c$ feature centres, which are evenly spaced values 
along the badge size interval ([0,1]) where the updates are focused. 
For the sake of simplicity we keep the location of these feature centres 
constant and  the same for all individuals; they are stored in 
vector $\symbf{b}$ . Each of these feature centres 
is associated with a weight for reward estimation and tendency to play 
retreat in a given interaction. The reward estimation and the tendency 
to retreat are calculated (in every interaction) as the sum of the weights 
associated with each feature centre, weighted by the response 
triggered by the feature (Fig. \ref{fig:model-struc} B, dots 
represent the feature weights). The response of each feature centre 
diminishes as a Gaussian function with the distance between the 
feature centre and the badge size of 
the partner (Fig. \ref{fig:model-struc} B, grey line). Formally, the 
reward estimation $\hat{R}$ when the focal individual ($i$) faces 
individual $j$ is given by,
\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_i-b_z|}{2\theta^2})} 
\end{equation}

@@

```{r model-struc,echo=FALSE,warning=FALSE,fig.width=8,fig.align='center',fig.height=7,fig.show='hold',fig.cap= "Model of communication in the context of aggressive interactions. In A, the reaction norm determines the badge size, *e.g* the size of a black band on a bird's chest. Red shows an informative reaction norm, whereas blue shows a uninformative reaction norm. B individuals perceiving the signal have a behavioral reaction norm that determines their probability of retreating (black line). The reaction norm arises from generalizing the information from the feature weights (black dots). Feature weights are values associated with a particular badge size and are updated as individuals interact with each other. The generalization of the information associated with each feature weight is represented by the grey line and its axis, which shows that the response triggered by the fourth feature diminishes as the value evaluated is further from the feature center. The learning process moves the feature weights (black dots) up or down depending on whether this leads to an increase in the estimated reward. Panels C and D show the effect of learning on the receiver strategy. Receivers in C face signallers with uninformative reaction norms (blue line in A). The uninformative reaction norm yields a badge size of 0.5, thus the behavioural reaction norm in C is only updated around that value. D Receivers face signallers with informative reaction norms (red line in A). Accordingly, receivers develop a threashold-like reaction norm where the decision to retreat increases with the badge size of the interacting partner. Colour scale in C and D indicates the quality of the individual. "}

@@

We first present the outcome of simulations where we prevented the 
evolution of the badge size (by setting the mutation rate to $0$), 
and assumed that all individuals in the population display either 
an uninformative or informative badge size with respect to their quality. 
These simulations show what type of receiver strategy develops through 
a learning process in such a monomorphic population. 
Figure \ref{fig:model-struc} (C and D) shows 
the receiver strategy developed through learning; panel C is for 
receivers that faced uninformative signals, and D for those facing informative ones. 
The simulations reveal that when learners face uninformative signals (Fig. \ref{fig:model-struc} C), they modify their probability of retreating 
depending on their own quality. 
Individuals with higher quality (red tones) have a 
high probability of attacking after the learning process; 
while individuals of lower quality 
(blue tones) mostly retreat from confrontations; in both cases 
individuals behave equally irrespective of the quality of
their opponent. Thus, learning splits the population of receivers into 
the two classic pure strategies of hawks and doves. 
Given that we assumed a monomorphic population 
with unresponsive reaction norms on the signalling side, the changes 
triggered by learning only affect a small range of badge sizes. Specifically, 
all learning occurred around a badge size of $0.5$, because all
badges in the population are of this size (panel C). 
In contrast, when receivers face informative reaction norms 
on the side of the signaller (Fig. \ref{fig:model-struc} B), 
receivers use information on the badge size of their interacting 
partners to determine whether to retreat or attack. 
The relationship is given by a threshold-like reaction norm, 
where the decision to retreat or to attack depends on the quality of the receiver. 
As expected, the higher the quality of the receiver, the larger the badge size 
of the signaller that triggers a retreat.

@@

When we allowed the badge size to evolve ($\alpha$ and $\beta$ 
changed subject to natural selection and genetic drift) and the signal
worked as a handicap (*i.e.* the cost of the badge was inversely 
proportional to the quality of the individual), the sender code evolved often
to produce an honest signal (Fig. \ref{fig:handicap}). The evolution of 
the sender code did not happen immediately after the start of the 
evolutionary process. Instead, the evolutionary dynamics of the reaction 
norm parameters (Fig. \ref{fig:handicap}, A) appeared to involve a set of 
steps. First, the intercept of the reaction norm ($\alpha$) 
describing the relationship between individual quality and badge size evolved to 
higher values, reducing the average badge size in the populations. 
This is because reaction norms segregating in the population are 
flat at the beginning of the simulation; consequently, 
the receiver code does not respond to the badge size 
and larger badges are costlier and do not trigger lower attack 
probabilities. During the first generations, the slope of the sender 
reaction norm ($\beta$)  remains close to the initial value of zero. 
At around generation 4000, the slope evolves toward positive values. 
Positive values in the slope imply that larger badges correlate 
with higher quality (Fig. \ref{fig:handicap} B). Receivers then learn to react to such 
correlation, increasing the probability of retreating when facing 
individuals with larger badges (Fig. \ref{fig:handicap} C). Hence, natural selection 
favours larger values of $\beta$, eventually leading to an evolutionary 
equilibrium in which badge size is an honest signal of quality mediated
by the learned responses of receivers (Fig. \ref{fig:handicap} B.4 and C.4).

@@

The amount of information that agents are able to collect through learning 
over their life time can strongly change the outcome of 
evolutionary dynamics. In the simulations presented in figure 
\ref{fig:handicap} and \ref{fig:noBadge}, 
individuals learned with high speed ($\mathrm{A} = 0.4$) and interacted 
repeatedly along their lifetime (2000 interactions on average). When we reduced 
the number of interactions that individuals experienced over their lifetime to 300 
on average, we saw a drastic increase in the phenotypic and genetic
variation present in the population. Populations start monomorphic 
with a value of zero for both the intercept and
the slope of the reaction norm, and mutations quickly build up a 
normal distribution around the starting 
value. Within the first 2000 generations, the unimodal distribution in 
the intercept ($\alpha$) splits into a bimodal one. 
Later in evolutionary time one of the peaks splits further, 
so at the end of the evolutionary simulation the distribution of the intercept 
($\alpha$) in the population shows three distinct peaks. In the case of $\beta$, the 
peaks in the distribution are not so clear-cut, but the variance of the 
distribution increases over time. These changes in the 
distribution of the parameters of the sender reaction norm 
imply that individuals can generally be classified into three 
distinct types (Fig. \ref{fig:branching}). Two types express a flat reaction 
norm with extreme values for the badge size, meaning that their badge size 
is not informative of their quality, whereas the third type shows 
intermediate badge sizes determined by individual quality 
(Fig. \ref{fig:branching} B.3-4). 
In this simulation, the values of the slope in the informative group
are negative, which implies that badge size correlates negatively with quality. 
Furthermore, the receiver reaction norms developed through learning, particularly those 
of individuals of intermediate quality, respond to the signal of their 
sender type by increasing the probability of retreating from a fight with 
individuals with smaller badges (Fig. \ref{fig:branching}).

@@

A limited number of interactions has an effect on the amount of variation in
the evolving parameters when the signal follows the handicap
principle as well. In fig \ref{fig:honestDiv}, we show simulations where individuals 
have on average 300 interactions in their life and the cost of the signal
is proportional to quality, following the handicap principle. 
The evolutionary process leads to a combination of reaction norm parameters where 
there is a positive correlation between the size of the badge and the quality 
of the individual. This relationship, however, is muddled by the fact that there 
are two clusters of values for the intercept ($\alpha$) and slope ($\beta$) 
of the reaction norm in the population. Thus, there are two types 
of reaction norms. One of the types has a steeper slope compared to the other. 
This effect of an increased variance in the trait distribution
is not only triggered by lower number of interactions. 
Larger variances are also found when we assume 
a lower speed of learning (data not shown). This suggest
that limits to the amount of information that individuals acquire through
learning allows the coexistence of different communication strategies within 
a population, something previously shown by @botero_Evolution_2010.

@@

The action of natural selection and genetic drift on the learning parameters 
do not override the conditions for the evolutionary branching 
[@dieckmann_Dynamical_1996] described earlier. 
Given how crucial naive behaviour is for the branching event 
described earlier, we tested whether evolutionary processes could lead 
key parameters of the learning process toward
values that prevented the genetic variation to build up. 
Figure \ref{fig:learnEvol} shows the dynamics from a set of simulations where we 
allow for the coevolution of the reaction norm as well as key parameters of
the learning module, namely the speed of learning ($A$) and the behavioural 
tendency expressed by individuals prior to learning. Top panels in 
figure \ref{fig:learnEvol} show that there is diversification process in this 
coevolutionary scenario as well. That is because, as is shown in the middle 
panels (B) of the same figure, the learning parameters stay within the range 
necessary for the branching event to take place. The speed of learning 
(Fig. \ref{fig:learnEvol} B.1) stays well above zero throughout the simulation. 
In contrast, the initial behavioral tendency maintain values around zero. 
Thus, evolutionary processes acting on the learning parameters 
did not override the conditions necessary for frequency-dependent selection to boost 
genetic variation.
